---
title: "4_variance"
author: "Nick Schultz"
date: "3/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
ozone <- read.csv("gardens.csv")

attach(ozone)

ozone
```
## calculate variance aka sum of squares
```{r}
mean(gardenA)

gardenA - mean(gardenA)

(gardenA - mean(gardenA))^2

sum((gardenA - mean(gardenA))^2)

sum((gardenA - mean(gardenA))^2) / 9

sum((gardenB - mean(gardenB))^2) / 9

sum((gardenC - mean(gardenC))^2) / 9
```

## F test
are variances significantly different?
```{r}
var(gardenC)/var(gardenB)

2 * (1 - pf(q = 10.667, df1 = 9, df2 = 9))

var.test(gardenB, gardenC)
```
Constancy of variance is assumed when carry out standard tests, i.e. Student's t test, to compare two means, or ANOVA to compare three or more means.
```{r}
mean(gardenA) == mean(gardenB)
var(gardenA) == var(gardenB)
```



When the variances are different, do not compare the means.
```{r}
mean(gardenB) == mean(gardenC)

(var(gardenB) == var(gardenC))

gardenB

gardenC
```

variance as a measure of unreliabiilty, i.e. standard error of the mean
```{r}
sqrt(var(gardenA)/10)
sqrt(var(gardenB)/10)
sqrt(var(gardenC)/10)
```
"The mean ozone concentration was 3.0 +- 0.365 pphm (1 s.e. = 10)" 

Confidence Intervals
If sample size n < 30, i.e. too small to use normal distribution, then use Student's t distribution.
```{r}
qt(p = 0.025, df = 9)
qt(p = .975, df = 9)

qt(p = .995, df = 9)
qt(.9975, 9)
```
The values of Student's t are number of standard errors to be expected with specified probability and for given number of degrees of freedom.

Confidence interval of mean based on small sample
```{r}
qt(.975, 9) * sqrt(1.33333/10)
```
"The mean ozone concentration in garden B was 5.0 +- 0.826 (95% CI, n = 10)."

Bootstrapping
```{r}
data <- read.csv("skewdata.csv")
attach(data)
names(data)
```

```{r}
plot(c(0, 30), c(0, 60), type = "n", xlab = "Sample size", 
     ylab = "Confidence interval")
for (k in seq(from = 5, to = 30, by = 3)) {
  a <- numeric(10000)
  for (i in 1:10000) {
    a[i] <- mean(sample(x = values, size = k, replace = TRUE))
  }
points(c(k,k), quantile(a, c(0.25, .975)), type = "b", pch = 21, bg = "red")
}

quantile(a, c(0.025, 0.975))

xv <- seq(5, 30, 0.1)
yv <- mean(values) + 1.96 * sqrt(var(values)/xv)
lines(xv, yv, col = "blue")
yv <- mean(values) - 1.96 * sqrt(var(values)/xv)
lines(xv, yv, col = "blue")

yv <- mean(values) + qt(.975, xv - 1) * sqrt(var(values)/xv)
lines(xv, yv, lty = 2, col = "green")
yv <- mean(values) - qt(.975, xv - 1) * sqrt(var(values)/xv)
lines(xv, yv, lty = 2, col = "green")

```

