---
title: "5_single_samples"
author: "Nick Schultz"
date: "3/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Single Samples

Questions:  1. what is mean value? 
            2. is mean value significantly different from current expectation or theory?
            3. what is the level of uncertainty associated with our estimate of the mean values?
            
First, we must establish:
            1. are the values normally distributed?
            2. are there outliers in the data?
            3. if data were collected over a period of time, is there evidence for serial correlation?
Non-normality, outliers, and serial correlation can invalidate inferences made by parametric tests, i.e. Student's t test. (Might need to use non-parametric techniques such as Wilcoxon's signed-rank test; if serial correlation, need to use time series analysis or mixed effectes model.)

```{r}
library(ggplot2)
library(readr)

data <- read.csv("example.csv")
data2 <- read_csv("example.csv")

head(data2)
names(data)

summary(data)
summary(data2)
boxplot(data$y)

ggplot(data = data2, aes(y)) +
  geom_boxplot()

hist(data$y)

ggplot(data, aes(y)) +
  geom_histogram(bins = 12, color = "white")
```

Calculations using z of the Normal distribution
suppose we have heights of 100 people; mean height = 170 cm; sd = 8 cm.
```{r}
ht <- seq(150, 190, 0.01)

plot(ht, dnorm(x = ht, mean = 170, sd = 8), type = "l", col = "brown",
     ylab = "Probability density", xlab = "Height")
```
We can ask three sorts of questions:
    1. shorter than particular height?
    2. taller than particular height?
    3. between one specified height and another?

Convert height into a value of z, i.e. convert height into a number of standard deviations from the mean.

Question:  what is the probability that one person selected at random will be less than 160 cm tall?

```{r}
z <- (160 - 170) / 8

pnorm(z)
```
What is probabilty that person is taller than 185 cm?
```{r}
z2 <- (185 - 170) / 8

1 - pnorm(z2)
```

What is probability that person is between 165 cm and 180 cm?
```{r}
z_1 <- (165 - 170) / 8
z_2 <- (180 - 170) / 8

pnorm(z_2) - pnorm(z_1)
```

Plots for testing normality of samples
```{r}
data <- read.csv("skewdata.csv")

qqnorm(data$values)
qqline(data$values, lty = 2)
```

Data with outliers
```{r}
light <- read.csv("light.csv")

hist(light$speed)
summary(light$speed)
```
Outlier rule of thumb: value 1.5 times the interquartile range
```{r}
850 - ((980 - 850) * 1.5)
980 + ((980 - 850) * 1.5)
```

Inference in one-sample case
```{r}
wilcox.test(x = light$speed, mu = 990)
```

Bootstrap in hypothesis testing with single samples
Question: how likely is it that the population mean we are trying to estimate with our random sample of 100 values is as big as 990?

  1. Take 10,000 random samples with replacement using n = 100 from 100 values of ligth and calculate 10,000 values of the mean.
  2. What is probability of obtaining a mean as large as 990?
```{r}
a <- numeric(10000)
for (i in 1:10000) a[i] <- mean(sample(x = light$speed, replace = TRUE))
hist(a)

max(a)
mean(a)
sd(a)

```

Skew

```{r}
skew <- function(x) {
  m3 <- sum((x - mean(x))^3)/ length(x)
  s3 <- sqrt(var(x))^3
  m3/s3
}

hist(data$values, main = "", col = "green")

skew(data$values)

t_test <- skew(data$values)/sqrt(6/length(data$values))

1 - pt(t_test, length(data$values) - 2)

skew(sqrt(data$values))/sqrt(6/length(data$values))
skew(log(data$values))/sqrt(6/length(data$values))

```

Kurtosis
```{r}
kurtosis <- function(x){
  m4 <- sum((x - mean(x))^4)/length(x)
  s4 <- var(x)^2
  m4/s4 - 3
}

kurtosis(data$values)

kurtosis(data$values)/sqrt(24/length(data$values))
```

